---
title: "Data visualization in R with ggplot2"
output:
  pdf_document: default
  html_document:
    df_print: paged
---
  
## Lesson preamble
  
> ### Learning Objectives
>
> - Create standard plots with `ggplot2` (scatter, line, hist)
> - Learn to modify plot appearance with `theme()`
> - Faceting plots
> - Composing multiple plots with `patchwork`
> - Annotating plots (model fits + text)
>
> ### Setup/Required packages
>
> - `install.packages('tidyverse')`
> - `install.packages('patchwork')`
>


```{r, echo=FALSE}
knitr::opts_chunk$set(eval=FALSE)
```
-----
  
# Plotting with ggplot2

## Introduction

`ggplot2` is a plotting package that makes it simple to create complex plots
from data frames. The name `ggplot2` comes from its inspiration, the book "A
grammar of graphics", and the main goal is to allow coders to express their
desired outcome on a high level instead of telling the computer every detail
about what will happen. For example, you would say "color my data by species"
instead of "go through this data frame and plot any observations of species1 in
blue, any observations of species2 in red, and so on". Thanks to this
functional way of interfacing with data, only minimal changes are required if
the underlying data change. This helps in thinking about the data and creating
publication quality plots with minimal amounts of adjustments and tweaking.

`ggplot2` graphics are built step by step by adding new elements, or layers.
Adding layers in this fashion allows for extensive flexibility and
customization of plots. 

```{r}
library(ggplot2)
library(dplyr) # for some helper functions

# alternatively, load in the full tidyverse
# library(tidyverse)
```


## The `diamonds` dataset

We will be working with the `diamonds` dataset in this lesson. This is a large
(~54,000 x 10) dataset containing measurements and prices of various diamonds.
If you have an older computer and would like to just work with a subset of the
data, you can create that subsetted data frame as follows:
  
```{r}
d <- sample_frac(diamonds, size = 0.1) # should be 5400 rows instead
```

This will create a new data frame `d` containing 10% of the rows in the original
dataset following random sampling.

## Plotting with `ggplot2` - the geoms

### Making our first scatter plot

To build a ggplot, we need to:
  
1. Use the `ggplot()` function and bind the plot to a specific data frame using
the `data` argument

```{r}
ggplot(data = diamonds)
```

If the arguments are provided in the right order then the
names of the arguments can be omitted

```{r}
ggplot(diamonds)
```

2. Define aesthetics (`aes`), by selecting the variables to be plotted and the
variables to define the presentation such as plotting size, shape color, etc.

```{r}
ggplot(diamonds, aes(x = carat, y = price))
```

3. Add `geoms` -- geometrical objects as a graphical representation of the data
in the plot (points, lines, bars). `ggplot2` offers many different geoms; we
will use a few common ones today, including:
* `geom_point()` for scatter plots, dot plots, etc.
* `geom_histogram()` for histograms

To add a geom to the plot, use the `+` operator. Because we have two continuous
variables, let's use `geom_point()` first:

```{r}
ggplot(diamonds, aes(x = carat, y = price)) +
   geom_point()
```

The `+` in the `ggplot2` package is particularly useful because it allows you
to modify existing `ggplot` objects. This means you can easily set up plot
"templates" and conveniently explore different types of plots, so the above
plot can also be generated with code like this:

```{r, first-ggplot-with-plus}
# assign plot to a variable
diamonds_plot <- ggplot(diamonds, aes(x = carat, y = price))

# draw the plot
diamonds_plot + geom_point()
```

Notes:

- Anything you put in the `ggplot()` function can be seen by any geom layers
  that you add (i.e., these are universal plot settings). This includes the x
and y axis you set up in `aes()`.
- You can also specify aesthetics for a given geom independently of the
  aesthetics defined globally in the `ggplot()` function.
- The `+` sign used to add layers must be placed at the end of each line
  containing a layer. If, instead, the `+` sign is added in the line before the
other layer, `ggplot2` will not add the new layer and R will return an error
message.

```{r, eval=FALSE}
# this will raise an error
ggplot(diamonds, aes(x = carat, y = price)) 
+ geom_point()
```


### Building plots iteratively + scatter plots

Building plots with `ggplot2` is typically an iterative process. We start by
defining the dataset we'll use, lay the axes, and choose a geom:
  
```{r}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point()
```

Then, we start modifying this plot to extract more information from it. For
instance, we can add transparency (`alpha`) to reduce overplotting:
  
```{r}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point(alpha = 0.5)
```

From the eye test, we can see some evidence of clusters in this scatter plot.
It's possible that one of the categorical variables we have in the data could
explain this pattern. Coloring the data points according to a categorical
variable is an easy way to assess if there seems to be correlation. Let's try
this with `cut`.

```{r}
ggplot(diamonds, aes(x = carat, y = price, color = cut)) +
  geom_point(alpha = 0.5)
```

While some clustering is visible here, the pattern is not especially clear.
Perhaps something else in our dataset might be a better predictor. Let's try
`clarity` instead:

```{r}
ggplot(diamonds, aes(x = carat, y = price, color = clarity)) +
geom_point(alpha = 0.5)
```

Turns out clarity appears to be a strong predictor here after all.  Given that
IF1 is the best possible clarity measurement, we see that even 1-2 carat IF1
diamonds fetch equivalent prices to 4-5 carat I1 (lowest clarity) diamonds.

### Histograms

Similarly to how we made the above scatter plot, histograms in `ggplot2` can
simply be tacked on as another geom. The difference is that histograms do not
take in a y aesthetic, since that'll be 'count' by definition:

```{r}
ggplot(diamonds, aes(x = carat)) +
  geom_histogram()
```

The binning of a given histogram can be modified by either specifying a
binwidth or a fixed number of bins:

```{r}
ggplot(diamonds, aes(x = carat)) +
  geom_histogram(binwidth = 0.05)
```

## Model fits with `geom_smooth`

Fitting models to data is a common means of investigating whether certain
relationships are at play in a given dataset. `ggplot2` makes visualizing
these easy by largely consolidating line-fitting into a single geom -
`geom_smooth`.

By default, `geom_smooth` will automatically select what it thinks is the best
model for your dataset. However, rarely will it select a linear fit. To draw a
linear fit (or any other sort of fit, i.e. LOESS, quadratic, etc.) we have to
specify it:

```{r}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = 'lm')
```

Adding the color aesthetic we had previously will cause lines to be fit to each individual group:

```{r}
ggplot(diamonds, aes(x = carat, y = price, color = clarity)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = 'lm')
```

To still fit a line across the entire dataset, specify the color aesthetic _within_ `geom_point`, and not globally.

```{r}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point(aes(color = clarity), alpha = 0.5) +
  geom_smooth(method = 'lm')
```

## Intermediate `ggplot` 

### Customizing plot appearance

Although our plots have been rendering just fine so far, we may want to go a
step further and personalize the plots in some way. Fortunately, `ggplot2`
features a whole host of customization options, all nestled within a handy
function called `theme`. Within `theme`, we can specify certain plot elements
as named arguments and then use what are called the `element` family of
functions to style them.

Let's start by regenerating the original scatter plot:

```{r}
p <- ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point(aes(color = clarity), alpha = 0.5) # important to specify aes in geom_point for later in the lesson
p
```

To style text-based elements, such as axis labels as well as axis tick labels,
we would use the `element_text` helper function:

```{r}
p +
  theme(axis.title = element_text(family = 'Helvetica', size = 14), 
        axis.text = element_text(family = 'Helvetica', size = 14, color = 'black'))
```

To style lines, such as axis lines, the `element_line` helper function is the
way to go:

```{r}
p +
  theme(axis.title = element_text(family = 'Helvetica', size = 14), 
        axis.text = element_text(family = 'Helvetica', size = 14, color = 'black'),
        axis.line = element_line(color = 'black', size = 0.75))
```

But this makes the grey background stick out more. We can get rid of an element
altogether by setting it to `element_blank()`.

```{r}
p +
  theme(axis.title = element_text(family = 'Helvetica', size = 14), 
        axis.text = element_text(family = 'Helvetica', size = 14, color = 'black'),
        axis.line = element_line(color = 'black', size = 0.75),
        panel.background = element_blank())
```

Some axis elements also have specific `x` and `y` versions; for instance,
although we've modified `axis.text` with a single argument above, we can
also specify certain settings that specifically apply to `axis.text.x`. Let's
have the labels on a 45 degree angle:

```{r}
p +
  theme(axis.title = element_text(family = 'Helvetica', size = 14), 
        axis.text = element_text(family = 'Helvetica', size = 14, color = 'black'),
        axis.line = element_line(color = 'black', size = 0.75),
        panel.background = element_blank(),
        axis.text.x = element_text(angle = 45))
```

The final element type is `element_rect`. We can use this with
`panel.background` provide the plot with a light grey background:

```{r}
p +
  theme(axis.title = element_text(family = 'Helvetica', size = 14), 
        axis.text = element_text(family = 'Helvetica', size = 14, color = 'black'),
        axis.line = element_line(color = 'black', size = 0.75),
        panel.background = element_rect(color = 'light grey'))
```


`ggplot2` also comes preloaded with a host of pre-made themes - try them out:

```{r}
p + theme_bw()
p + theme_classic()
```

However, it's much more likely you'll want to create a theme specific
to your own needs. Fortunately, saving custom themes is quite straightforward.
Let's save the above theme to an object:

```{r}
fig_theme <- theme(axis.title = element_text(family = 'Helvetica', size = 14), 
        axis.text = element_text(family = 'Helvetica', size = 14, color = 'black'),
        axis.line = element_line(color = 'black', size = 0.75),
        panel.background = element_blank())
```

This can now be added to plots like any other `ggplot` object:

```{r}
p +
  fig_theme
```

We'll be revisiting another use of this concept later in the lesson.

### Legend formatting

Legend formatting involves a few extra considerations. First of all,
if you want to remove a legend altogether, this is done with `guides()`:

```{r}
p +
  fig_theme +
  guides(color = FALSE)
```

Notice that the argument given the `guides` is the aesthetic (in this
case, color) that we were using to generate the legend. 

Let's move our legend around and style the text a little:

```{r}
p +
  fig_theme +
  theme(legend.position = 'bottom', # special arg - values: 'top', 'left', etc
        legend.title = element_text(size = 12),
        legend.text = element_text(size = 12))
```

`legend.key` and `legend.background` also take in `element_rect`. Note
the difference between `color` (border color) and `fill` (fill color):

```{r}
p +
  fig_theme +
  theme(legend.position = 'bottom',
        legend.title = element_text(size = 12),
        legend.text = element_text(size = 12),
        legend.key = element_rect(color = 'black', fill = 'white'),
        legend.background = element_rect(color = 'black'))
```

These legend settings can also be saved as an object for reuse. We'll set `legend.position` back to `right`.

```{r}
legend_theme <- theme(legend.position = 'right',
        legend.title = element_text(size = 12),
        legend.text = element_text(size = 12),
        legend.key = element_rect(color = 'black', fill = 'white'),
        legend.background = element_rect(color = 'black'))
```



## Plot annotation

### The `labs()` function

The `labs` function is a useful catch-all to rename your axes and more.
Let's add custom x and y axis titles:

```{r}
p + 
  fig_theme +
  legend_theme +
  labs(x = 'Carat',
       y = 'Price ($)')
```

Some axis titles require mathematical formatting. Let's pretend,
for the sake of this example, that carats were abbreviated to
'c^2', and that this is what we wanted the axis title to be.
Mathematical expressions can be created using the `expression` helper
function. Note that inputs to `expression` are *not* in quotes:

```{r}
p +
  fig_theme +
  legend_theme +
  labs(x = expression(c^{2}),
       y = 'Price ($)')
```

`expression` works off of syntax known as **plotmath** - plotmath syntax is why
we need the `{}` around our superscript. We won't go into further details here,
but a full syntax reference can be found by running `?plotmath`.

To combine plotmath with text, use `paste()` within `expression`. Be
mindful of including spaces where needed; `expression` is quite literal.

```{r}
p +
  fig_theme +
  legend_theme +
  labs(x = expression(paste('Values (', c^{2}, ')')),
       y = 'Price ($)')
```

### Adding text directly to a plot

Let's say we fit a linear model with carat as the predictor and price as the
response. How can we add the R squared and p-value of our model to the plot
within R?

`ggplot` offers a useful `annotate` function for this reason. First, let's
quickly fit this model and get our R squared and model p-value:

```{r}
lm(price ~ carat, data = diamonds) %>% 
  summary()
```

`annotate` uses positioning values based off of the axis scales of the plot.
Let's add the R squared to the bottom right:

```{r}
p +
  fig_theme +
  legend_theme +
  geom_smooth(method = 'lm', color = 'black') +
  annotate('text', x = 3.5, y = 3000,
           label = 'R^2 = 0.8493')
```

We can style the added text with plotmath by setting `parse = TRUE`. Note that
we change the label slightly to match plotmath syntax.

```{r}
p +
  fig_theme +
  legend_theme +
  geom_smooth(method = 'lm', color = 'black') +
  annotate('text', x = 3.5, y = 3000,
           label = 'italic(R)^{2} == 0.8493', parse = TRUE)
```

Let's add the p-value:

```{r}
p +
  fig_theme +
  legend_theme +
  geom_smooth(method = 'lm', color = 'black') +
  annotate('text', x = 3.5, y = 3000,
           label = 'italic(R)^{2} == 0.8493', parse = TRUE) +
  annotate('text', x = 3.53, y = 500,
           label = 'p < 2.2 %*% 10^{-16}', parse = TRUE)
```

### An aside - changing the dimensions of a plot

Notice how in the linear fit above, the line extends much further out. If we
wanted to change the 'zoom' of the plot to focus on the main cluster of points,
we can use `coord_cartesian`:

```{r}
p +
  fig_theme +
  legend_theme +
  geom_smooth(method = 'lm', color = 'black') +
  annotate('text', x = 3.5, y = 3000,
           label = 'italic(R)^{2} == 0.8493', parse = TRUE) +
  annotate('text', x = 3.53, y = 500,
           label = 'p < 2.2 %*% 10^{-16}', parse = TRUE) +
  coord_cartesian(x = c(0, 4), y = c(0, 20000))
```

This does not affect the points or the linear fit in any way; it just rescales the plot itself. 

## Making multiple plots

### Faceting

`ggplot` has a special technique called *faceting* that allows the user to
split one plot into multiple subplots based on a variable included in the
dataset. We will use `facet_wrap` to make the same scatter plot as above (price
by carat) but split it by clarity:

```{r}
ggplot(diamonds, aes(x = carat, y = price, color = clarity)) +
  geom_point(alpha = 0.5) +
  facet_wrap(~ clarity) +
  fig_theme
```

Alternatively, it may be useful to facet by two categorical variables. This is
where `facet_grid` comes in:

```{r}
ggplot(diamonds, aes(x = carat, y = price, color = clarity)) +
  geom_point(alpha = 0.5) +
  facet_grid(cut ~ clarity) +
  fig_theme
```

Notice how our facet labels don't match the rest of our theme. This is because
we did not specify anything for labels in creating `fig_theme` above. Facet
labels can be styled within `theme` using `strip.background` (`element_rect`)
and `strip.text` (`element_text`). Try it out! 

### Composing multiple plots with `patchwork`

While faceting is useful to make subplots with, it's specifically limited to
doing so off of categorical variables. What if we wanted to show both our
scatter plot above *and* a histogram of carat in the same figure?

This is where `patchwork` comes in. `patchwork` allows for entire plots
to be strung together with `+`, as if they were themselves geoms. Let's
prep our plots:

```{r}
fig_1a <- p +
  fig_theme +
  legend_theme +
  labs(x = 'Carat', y = 'Price', tag = 'A')

fig_1b <- ggplot(diamonds, aes(x = carat)) +
  geom_histogram(binwidth = 0.05) +
  fig_theme +
  labs(x = 'Carat', y = 'Count', tag = 'B')
```

Note that by saving our theme to `fig_theme`, we can ensure these
two plots have a consistent look. We've also added `tag` to `labs`
since these are now going to be subplots in a single figure.
Note that `tag` requires `ggplot2` 3.0 -- if you see an error,
you may need to update the package.

Let's put these together:

```{r}
library(patchwork)

fig_1a + fig_1b
```

It's really that simple! If we want these plot tags to be styled differently, we
would have to update `plot.tag` (`element_text`) within `theme`. But here we
see the advantage of consistent styling -- these plots look nice and
'standardized'!

We can modify plot layout using `plot_layout`, which is tacked on
with the `+`. 

```{r}
fig_1a + fig_1b + plot_layout(nrow = 2, ncol = 1)
```

Finally, the combined plot can be saved to an object:

```{r}
fig_1 <- fig_1a + fig_1b
```


# Saving plots

Plots are saved using the `ggsave` function. Let's save our
combined `fig_1` plot:

```{r}
ggsave('fig_1.png', plot = fig_1)
```


# Wrapping up

We've covered many of the core elements of `ggplot2` today:

* Basic structure of a ggplot call
* A few introductory `geoms`
    * `geom_point`
    * `geom_histogram`
* Model fits with `geom_smooth`
* Customizing plots with `theme`
    * Using `element` helper functions - `element_text` and `element_line`
* Annotating plots
* Faceting with `facet_wrap` and `facet_grid`
* Composing multiple plots with `patchwork`
* Saving plots with `ggsave`

## Further resources

* EEB313 - [Introduction to `dplyr` and `ggplot2`][eeb313]
* Data Carpentry's [Data visualization with `ggplot2`][dc-r]
* The [data visualization][data-vis] chapter from Hadley Wickham and Garrett Grolemund's _R for Data Science_
* Hadley Wickham's [_ggplot2_ book][ggplot-book]
* Winston Chang's [_R Graphics Cookbook_][cookbook]

[cookbook]: http://www.cookbook-r.com/Graphs/
[data-vis]: https://r4ds.had.co.nz/data-visualisation.html
[dc-r]: https://datacarpentry.org/R-ecology-lesson/04-visualization-ggplot2.html
[eeb313]: https://uoftcoders.github.io/rcourse/lec04-dplyr.html
[ggplot-book]: https://www.amazon.com/dp/0387981403/ref=cm_sw_su_dp?tag=ggplot2-20



